Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-24 paper/main.tex
reworked 'understanding' statement

The difference between ``understanding'' and ``memorizing,'' as framed
by researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08,
  Macl05}, has profound analogs in the fields of natural language
processing and natural language understanding. For example,
considering the raw contents of a document (e.g., its constituent
symbols, letters, and words) might provide some clues as to what the
document is about, just as memorizing a passage might provide some
ability to answer simple questions about it. However, text embedding
models~\citep[e.g.,][]{LandDuma97, DeerEtal90, BleiEtal03,
  BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20, ViswEtal17} also
attempt to capture the deeper meaning \textit{underlying} those atomic
elements. These models consider not only the co-occurrences of those
elements within and across documents, but (in many cases) also
patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical
properties of the elements, and other high-level characteristics of
how they are used~\citep{Mann20, Mann21a}. According to these models,
the deep conceptual meaning of a document may be captured by a feature
vector in a high-dimensional representation space, wherein nearby
vectors reflect conceptually related documents. A model that succeeds
at capturing an analogue of ``understanding'' is able to assign nearby
feature vectors to two conceptually related documents, \textit{even
  when the specific words contained in those documents have limited
  overlap}. In this way, ``concepts'' are defined implicitly by the
model's geometry~\citep[e.g., how the embedding coordinate of a given
word or document relates to the coordinates of other text
embeddings; ][]{PianHill22}.

>>>>

The difference between ``understanding'' and ``memorizing,'' as framed
by researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08,
  Macl05}, has profound analogs in the fields of natural language
processing and natural language understanding. For example,
considering the raw contents of a document (e.g., its constituent
symbols, letters, and words) might provide some clues as to what the
document is about, just as memorizing a passage might provide some
ability to answer simple questions about it. However, text embedding
models~\citep[e.g.,][]{LandDuma97, DeerEtal90, BleiEtal03,
  BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20, ViswEtal17} also
attempt to capture the deeper meaning \textit{underlying} those atomic
elements. These models consider not only the co-occurrences of those
elements within and across documents, but (in many cases) also
patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical
properties of the elements, and other high-level characteristics of
how they are used~\citep{Mann20, Mann21a}. 
To be clear, this is not to say that text embedding models themselves 
are capable of ``understanding'' deep conceptual meaning in any 
traditional sense. But rather, their ability to capture the underlying 
\textit{structure} of text documents beyond their surface-level contents 
provides a computational framework through which those document's 
deeper conceptual meaning may be quantified, explored, and understood.
According to these models,
the deep conceptual meaning of a document may be captured by a feature
vector in a high-dimensional representation space, wherein nearby
vectors reflect conceptually related documents. A model that succeeds
at capturing an analogue of ``understanding'' is able to assign nearby
feature vectors to two conceptually related documents, \textit{even
  when the specific words contained in those documents have limited
  overlap}. In this way, ``concepts'' are defined implicitly by the
model's geometry~\citep[e.g., how the embedding coordinate of a given
word or document relates to the coordinates of other text
embeddings; ][]{PianHill22}.
